---
title: "Reproducible Research - Peer Assessment 1"
author: "Carlos Ortega"
date: "Monday, August 11, 2014"
output:
  html_document:
    number_sections: yes
---

***
***

# Loading and preprocessing the data
## Get Data
```{r, echo=TRUE}
datIn <- read.csv("activity.csv", header=T)
head(datIn)
```

***

# What is mean total number of steps taken per day?
## Histogram number of steps each day
```{r, echo=TRUE, fig.width=9, fig.height=7}
library(lattice)
Histgr <- histogram(
                     ~ steps | date
                    ,data=datIn
                    ,xlab="Number of Steps"
                    ,ylab="Frequency"
                    ,main="Number of Steps per day"
                    ,par.strip.text=list(font=2, col="black",cex=0.75)
                    ,as.table=TRUE
                    )
print(Histgr)
```

## Mean and median of steps per day, working only with complete.cases
```{r, echo=TRUE}
library(sqldf)
DF <- datIn[ complete.cases(datIn),]
DFres <- sqldf("select date, avg(steps) as Mean, median(steps) as Median from DF group by date order by date")
head(DFres)
```
  
***

# What is the average daily activity pattern?
## Average daily activity pattern
```{r, echo=TRUE, fig.width=9, fig.height=7}
DFactivity <- sqldf("select interval, avg(steps) as Mean from DF group by interval order by interval")
head(DFactivity)
```
```{r, echo=TRUE, fig.width=9, fig.height=7}
plot(
     DFactivity$interval, DFactivity$Mean, type="l", col="blue", las=1
    ,xlab="Interval", ylab="Mean number of steps", main="Average Steps per Interval",
    ,font.axis=2, col.axis="blue", col.lab="red"
     )
grid()
```

## Interval with the maximum number of steps
```{r, echo=TRUE}
DFactivity$interval[which.max(DFactivity$Mean)]
```

***

# Imputing missing values
## Number of NAs
```{r, echo=TRUE}
numberNas <- dim(datIn)[1] - dim( datIn[complete.cases(datIn),])[1]
numberNas
```


## Filling NAs  - Filled with average for the interval acrross all days
```{r, echo=TRUE}
theNAs <- datIn[is.na(datIn),]
theNAsfilled <- merge(theNAs, DFactivity, by.x="interval", by.y="interval")
noNAs <- theNAsfilled[,c(4,3,1)]
names(noNAs) <- c('steps', 'date', 'interval')
datInNew <- rbind.data.frame(DF, noNAs)
```


## Histogram number of steps taken each day
```{r, echo=TRUE, fig.width=9, fig.height=7}
library(lattice)
HistgrNew <- histogram(
  ~ steps | date
  ,data=datInNew
  ,xlab="Number of Steps"
  ,ylab="Frequency"
  ,main="Number of Steps per day (filled NAs)"
  ,par.strip.text=list(font=2, col="black",cex=0.75)
  ,as.table=TRUE
)
print(HistgrNew)
```

## Mean and median of steps per day, working only with complete.cases
```{r, echo=TRUE}
library(sqldf)
DFresnoNA <- sqldf("select date, avg(steps) as Mean, median(steps) as Median from datInNew group by date order by date")
head(DFresnoNA)
```

## Impact due to filled NAs
```{r, echo=TRUE}
DFscompa <- merge(DFres, DFresnoNA, by.x='date', by.y='date')
summary(DFscompa$Mean.x)
summary(DFscompa$Mean.y)
all.equal(DFscompa$Mean.x, DFscompa$Mean.y)
```
**No difference. NAs affect only to complete days.**

***

# Are there differences in activity patterns between weekdays and weekends?
## Difference activity between weekday and weekend
```{r, echo=TRUE}
library(lubridate)
datInNew$wday <- wday(ymd(datInNew$date))
datInNew$typeday <- ifelse(datInNew$wday==1 | datInNew$wday==6,'weekend','weekday')
head(datInNew)
```

(Sunday = 1 / Saturday = 6)

## Averaged by interval and typeday
```{r, echo=TRUE}
DFinterday <- sqldf("select avg(steps) as Mean, interval, typeday from datInNew group by interval, typeday")
head(DFinterday)
```

## Plot steps between weekend - weekday
```{r, echo=TRUE, fig.width=9, fig.height=7}
library(lattice)
xyweek <- xyplot(
                  Mean ~ interval | typeday
                 ,data=DFinterday
                 ,type="l"
                 ,xlab=list("Interval",font=2,col="blue",cex=1.3)
                 ,ylab=list("# of steps",font=2,col="blue",cex=1.3)
                 ,scales=list(cex=1, font=2, col="black")
                 ,par.strip.text=list(font=2, col="black",cex=1.25)
                 ,panel=function(x,y,...) {
                   panel.grid(h=-1, v=-1, lty=2) 
                   panel.xyplot(x,y,...) }
                 ,layout=c(1,2) 
                )
print(xyweek)
```




